### 2025-03-15 paper/quarto_ecml/paper.tex
table fixed and I need a break after this latex nightmare
vamooooos

We have simulated this scenario using synthetic data with independent
\emph{age} and \emph{debt}, and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels show the outcomes for different
training procedures using the same model architectures (a linear
classifier). In panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we used CT. Only in panels
(c) and (d) do we impose the mutability constraint on \emph{age} at test
time. In each case, we show the decision boundary (in green) and the
training data colored according to their ground-truth label: orange
points belong to the target class, \(y^+=2\), blue points belong to the
non-target class, \(y^-=1\). Stars indicate CEs in the target class
generated at test time using generic gradient descent until convergence.

In all cases the counterfactuals are all valid but their quality
differs. In panel (a), they are not plausible: they do not comply with
the distribution of the factuals in \(y^+\) to the point where they form
a clearly distinguishable cluster. In panel (b), they are highly
plausible, meeting the first objective of Def. \ref{def-explainability}.
In panel (c), the CEs involve substantial reductions in \emph{debt} for
younger applicants. By comparison, counterfactual paths are shorter on
average in panel (d) where we have protected the immutable \emph{age} as
described in Section~\ref{sec-constraints}. Due to the classifier's
lower sensitivity to \emph{age}, recommendations with respect to
\emph{debt} are much more homogenous and do not unfairly punish younger
individuals. These counterfactuals are also plausible with respect to
the mutable feature. Thus, we consider the model in panel (d) as the
most explainable according to Def. \ref{def-explainability}.

@@

\caption{\label{fig-poc}Illustration of how CT improves model
explainability.}

@@

Our key outcome of interest is how well models perform with respect to
explainability (Def. \ref{def-explainability}). To this end, we focus
primarily on the plausibility and cost of faithfully generated
counterfactuals at test time. To measure the cost of counterfactuals, we
follow the standard convention of using distances (\(\ell_1\)-norm)
between factuals and counterfactuals as a proxy. For plausibility, we
assess how similar counterfactuals are to observed samples in the target
domain, \(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the
distance-based metric used in \cite{altmeyer2024faithful},

@@

In addition to cost and plausibility, we also compute other standard
metrics to evaluate counterfactuals at test time including validity and
redundancy. Finally, we also assess the predictive performance of models
using standard metrics.

@@

@@

\subsubsection{Plausibility.}\label{sec-plaus}

Table~\ref{tbl-main} presents our main empirical findings. The top five
rows show the percentage reduction in implausibility according to
Equation~\ref{eq-impl-dist} for varying degrees of the energy penalty
used for \emph{ECCo} at test time. Row 6 shows the reduction in
implausibility as measured by Equation~\ref{eq-impl-div} and aggregated
across all test specifications of \emph{ECCo}. Rows 7 and 8 show the
test accuracies for the model trained with CT and conventionally trained
models (``vanilla'').

For all datasets except \emph{OL} and across all test settings, the
average distance of CEs from observed samples in the target class is
reduced, indicating improved plausibility. The magnitude of improvements
varies. For the simple synthetic datasets, distance reductions range
from around 20-40\% (\emph{LS}, \emph{Moon}) to almost 60\%
(\emph{Circ}). For the real-world tabular datasets, improvements tend to
be smaller but still substantial, with around 10-15\% for \emph{CH},
11-28\% for \emph{GMSC}, 7-8\% for \emph{Cred}, and around 3\% for
\emph{Adult}. For our only vision dataset (\emph{MNIST}), distances are
reduced by up to 9\%. The results for our proposed divergence metric are
qualitatively similar, but generally even more pronounced: for the
\emph{Circ} dataset, implausibility is reduced by almost 94\% to
virtually zero as we verified by the absolute outcome. Improvements for
other datasets range from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For
\emph{OL} the reduction is negative, consistent with the distance-based
metric. \emph{MNIST} is the only dataset for which the two metrics
disagree.

We find that CT can substantially improve the quality of explanations
learned by models without sacrificing predictive performance. Test
accuracy for CT is virtually identical to the baseline for \emph{Adult},
\emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and even slightly
improved for \emph{Cred}. Exceptions to this general pattern are
\emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we observe reduction
in test accuracy of 2, 5, and 15 percentage points respectively. In this
context, we also note that we have not optimized the models (very small
networks) for predictive performance at all.

@@

\caption{\label{tbl-main}Key performance metrics across all datasets.
\textbf{Plausibility}: Rows 1-5 show the percentage reduction in
implausibility via Equation~\ref{eq-impl-dist} for varying degrees of
the energy penalty used for \emph{ECCo} at test time; row 6 shows the
reduction in implausibility via Equation~\ref{eq-impl-div} and
aggregated across all test specifications. \textbf{Accuracy} (rows 7-8):
test accuracies for models trained with CT and the baseline.
\textbf{Actionability} (row 9): average reduction in costs when imposing
mutability constraints.}

@@

\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
\input{tables/main.tex}
\end{tabular*}

@@

\subsubsection{Actionability.}\label{sec-act}

>>>>

We have simulated this scenario using synthetic data with independent
\emph{age} and \emph{debt}, and Gaussian class-conditional densities in
Figure~\ref{fig-poc}. The four panels show the outcomes for different
training procedures using the same model architectures (a linear
classifier). In panels (a) and (c) we have trained the models
conventionally, while in panels (b) and (d) we used CT.

In all cases, all counterfactuals (stars) are valid---they have cross
the decision boundary (green)---but their quality differs. In panel (a),
they are not plausible: they do not comply with the distribution of the
factuals in \(y^+\) to the point where they form a clearly
distinguishable cluster. In panel (b), they are highly plausible,
meeting the first objective of Def. \ref{def-explainability}. In panel
(c), the CEs involve substantial reductions in \emph{debt} for younger
applicants. By comparison, counterfactual paths are shorter on average
in panel (d) where we have protected the immutable \emph{age} as
described in Section~\ref{sec-constraints}. Due to the classifier's
lower sensitivity to \emph{age}, recommendations with respect to
\emph{debt} are much more homogenous and do not unfairly punish younger
individuals. These counterfactuals are also plausible with respect to
the mutable feature. Thus, we consider the model in panel (d) as the
most explainable according to Def. \ref{def-explainability}.

@@

\caption{\label{fig-poc}Illustration of how CT improves model
explainability: (a) conventional training, all mutable; (b) CT, all
mutable; (c) conventional, \emph{age} immutable; (d) CT, \emph{age}
immutable. The decision boundary is shown in green along with training
data colored according to their ground-truth label: \(y^+=2\) (orange)
and \(y^-=1\) (blue). Stars indicate CEs for the target class.}

@@

Our key outcome of interest is improved explainability (Def.
\ref{def-explainability}). To this end, we focus primarily on the
plausibility and cost of faithfully generated counterfactuals at test
time. To measure the cost of counterfactuals, we follow the standard
convention of using distances (\(\ell_1\)-norm) between factuals and
counterfactuals as a proxy. For plausibility, we assess how similar
counterfactuals are to observed samples in the target domain,
\(\mathbf{X}^+\subset\mathcal{X}^+\). We rely on the distance-based
metric used in \cite{altmeyer2024faithful},

@@

In addition to cost and plausibility, we compute other standard metrics
to evaluate counterfactuals including validity and redundancy. Finally,
we also assess the predictive performance of models using standard
metrics, including robust accuracy estimated on adversarially perturbed
data using FGSM \cite{goodfellow2014explaining}.

@@

To assess our proposed training regime, we investigate the improvements
in performance metrics when using CT on top of a weak baseline (BL):
specifically, a multilayer perceptron (\emph{MLP}). This is the best way
to get a clear picture of how effective CT is and consistent with how
assessment is done in the related literature
\cite{ross2021learning,teney2020learning,goodfellow2014explaining}.

@@

\subsubsection{Plausibility}\label{sec-plaus}

Table~\ref{tbl-main} presents our main empirical findings. For all
datasets except \emph{OL} and across all test settings, the average
distance of CEs from observed samples in the target class is reduced,
indicating improved plausibility. The magnitude of improvements varies.
For the simple synthetic datasets, distance reductions range from around
20-40\% (\emph{LS}, \emph{Moon}) to almost 60\% (\emph{Circ}). For the
real-world tabular datasets, improvements tend to be smaller but still
substantial, with around 10-15\% for \emph{CH}, 11-28\% for \emph{GMSC},
7-8\% for \emph{Cred}, and around 3\% for \emph{Adult}. For our only
vision dataset (\emph{MNIST}), distances are reduced by up to 9\%. The
results for our proposed divergence metric are qualitatively similar,
but generally even more pronounced: for the \emph{Circ} dataset,
implausibility is reduced by almost 94\% to virtually zero as we
verified by the absolute outcome. Improvements for other datasets range
from 28\% (\emph{Moon}) to 78\% (\emph{GMSC}). For \emph{OL} the
reduction is negative, consistent with the distance-based metric.
\emph{MNIST} is the only dataset for which the distance and divergence
metrics disagree. Upon visual inspection of the image counterfactuals we
find that CT clearly improves plausibility (see appendix).

\subsubsection{Predictive Performance}\label{sec-pred}

Test accuracy for CT is virtually identical to the baseline for
\emph{Adult}, \emph{Circ}, \emph{LS}, \emph{Moon}, and \emph{OL}, and
even slightly improved for \emph{Cred}. Exceptions to this general
pattern are \emph{MNIST}, \emph{CH}, and \emph{GMSC}, for which we
observe a reduction in test accuracy of 2, 5, and 15 percentage points
respectively. When looking at robust test accuracies (Acc.\(^*\)) for
these datasets in particular, we find that CT strongly outperforms the
baseline. In fact, we find that CT improves adversarial robustness on
all datasets.

@@

\caption{\label{tbl-main}Key performance metrics across all datasets.
\textbf{Plausibility}: Columns 2-6 show the percentage reduction in
implausibility (\(\text{IP}\)) for varying degrees of the energy penalty
used for \emph{ECCo} at test time; column 7 shows the reduction in
\(\text{IP}^*\) (\(\text{MMD}\)), aggregated across all test
specifications. \textbf{Accuracy} (columns 8-11): test accuracies and
robust accuracies (\(\text{Acc}^*\)) for CT and the baseline (BL).
\textbf{Actionability} (column 9): average reduction in costs when
imposing mutability constraints.}

@@

\input{tables/main.tex}

@@

\subsubsection{Actionability}\label{sec-act}

### 2025-03-15 paper/sections/experiments.qmd
table fixed and I need a break after this latex nightmare
vamooooos

Our key outcome of interest is how well models perform with respect to explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the distance-based metric used in @altmeyer2024faithful,

@@

In addition to cost and plausibility, we also compute other standard metrics to evaluate counterfactuals at test time including validity and redundancy. Finally, we also assess the predictive performance of models using standard metrics.

@@

@@

### Plausibility. {#sec-plaus}

@tbl-main presents our main empirical findings. The top five rows show the percentage reduction in implausibility according to @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time. Row 6 shows the reduction in implausibility as measured by @eq-impl-div and aggregated across all test specifications of *ECCo*. Rows 7 and 8 show the test accuracies for the model trained with CT and conventionally trained models ("vanilla").

For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies. For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the two metrics disagree.

We find that CT can substantially improve the quality of explanations learned by models without sacrificing predictive performance. Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon*, and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH*, and *GMSC*, for which we observe reduction in test accuracy of 2, 5, and 15 percentage points respectively. In this context, we also note that we have not optimized the models (very small networks) for predictive performance at all.

@@

```{=latex}
\begin{tabular*}{\linewidth}{@{\extracolsep{\fill}} cccccc }
\input{tables/main.tex}
\end{tabular*}
```

Key performance metrics across all datasets. **Plausibility**: Rows 1-5 show the percentage reduction in implausibility via @eq-impl-dist for varying degrees of the energy penalty used for *ECCo* at test time; row 6 shows the reduction in implausibility via @eq-impl-div and aggregated across all test specifications. **Accuracy** (rows 7-8): test accuracies for models trained with CT and the baseline. **Actionability** (row 9): average reduction in costs when imposing mutability constraints.
:::

### Actionability. {#sec-act}

>>>>

Our key outcome of interest is improved explainability (Def. \ref{def-explainability}). To this end, we focus primarily on the plausibility and cost of faithfully generated counterfactuals at test time. To measure the cost of counterfactuals, we follow the standard convention of using distances ($\ell_1$-norm) between factuals and counterfactuals as a proxy. For plausibility, we assess how similar counterfactuals are to observed samples in the target domain, $\mathbf{X}^+\subset\mathcal{X}^+$. We rely on the distance-based metric used in @altmeyer2024faithful,

@@

In addition to cost and plausibility, we compute other standard metrics to evaluate counterfactuals including validity and redundancy. Finally, we also assess the predictive performance of models using standard metrics, including robust accuracy estimated on adversarially perturbed data using FGSM [@goodfellow2014explaining].

@@

To assess our proposed training regime, we investigate the improvements in performance metrics when using CT on top of a weak baseline (BL): specifically, a multilayer perceptron (*MLP*). This is the best way to get a clear picture of how effective CT is and consistent with how assessment is done in the related literature [@ross2021learning;@teney2020learning;@goodfellow2014explaining].

@@

### Plausibility {#sec-plaus}

@tbl-main presents our main empirical findings. For all datasets except *OL* and across all test settings, the average distance of CEs from observed samples in the target class is reduced, indicating improved plausibility. The magnitude of improvements varies. For the simple synthetic datasets, distance reductions range from around 20-40% (*LS*, *Moon*) to almost 60% (*Circ*). For the real-world tabular datasets, improvements tend to be smaller but still substantial, with around 10-15% for *CH*, 11-28% for *GMSC*, 7-8% for *Cred*, and around 3% for *Adult*. For our only vision dataset (*MNIST*), distances are reduced by up to 9%. The results for our proposed divergence metric are qualitatively similar, but generally even more pronounced: for the *Circ* dataset, implausibility is reduced by almost 94% to virtually zero as we verified by the absolute outcome. Improvements for other datasets range from 28% (*Moon*) to 78% (*GMSC*). For *OL* the reduction is negative, consistent with the distance-based metric. *MNIST* is the only dataset for which the distance and divergence metrics disagree. Upon visual inspection of the image counterfactuals we find that CT clearly improves plausibility (see appendix).

### Predictive Performance {#sec-pred}

Test accuracy for CT is virtually identical to the baseline for *Adult*, *Circ*, *LS*, *Moon*, and *OL*, and even slightly improved for *Cred*. Exceptions to this general pattern are *MNIST*, *CH*, and *GMSC*, for which we observe a reduction in test accuracy of 2, 5, and 15 percentage points respectively. When looking at robust test accuracies (Acc.$^*$) for these datasets in particular, we find that CT strongly outperforms the baseline. In fact, we find that CT improves adversarial robustness on all datasets.

@@

```{=latex}
\input{tables/main.tex}
```

Key performance metrics across all datasets. **Plausibility**: Columns 2-6 show the percentage reduction in implausibility ($\text{IP}$) for varying degrees of the energy penalty used for *ECCo* at test time; column 7 shows the reduction in $\text{IP}^*$ ($\text{MMD}$), aggregated across all test specifications. **Accuracy** (columns 8-11): test accuracies and robust accuracies ($\text{Acc}^*$) for CT and the baseline (BL). **Actionability** (column 9): average reduction in costs when imposing mutability constraints.
:::

### Actionability {#sec-act}

### 2025-03-15 paper/sections/method.qmd
vamooooos

We have simulated this scenario using synthetic data with independent *age* and *debt*, and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT. Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time. In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence.

In all cases the counterfactuals are all valid but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. In panel (c), the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

![Illustration of how CT improves model explainability.](/paper/figures/poc.svg){#fig-poc}

>>>>

We have simulated this scenario using synthetic data with independent *age* and *debt*, and Gaussian class-conditional densities in @fig-poc. The four panels show the outcomes for different training procedures using the same model architectures (a linear classifier). In panels (a) and (c) we have trained the models conventionally, while in panels (b) and (d) we used CT.

<!-- Only in panels (c) and (d) do we impose the mutability constraint on *age* at test time. In each case, we show the decision boundary (in green) and the training data colored according to their ground-truth label: orange points belong to the target class, $y^+=2$, blue points belong to the non-target class, $y^-=1$. Stars indicate CEs in the target class generated at test time using generic gradient descent until convergence. -->

In all cases, all counterfactuals (stars) are valid---they have cross the decision boundary (green)---but their quality differs. In panel (a), they are not plausible: they do not comply with the distribution of the factuals in $y^+$ to the point where they form a clearly distinguishable cluster. In panel (b), they are highly plausible, meeting the first objective of Def. \ref{def-explainability}. In panel (c), the CEs involve substantial reductions in *debt* for younger applicants. By comparison, counterfactual paths are shorter on average in panel (d) where we have protected the immutable *age* as described in @sec-constraints. Due to the classifier's lower sensitivity to *age*, recommendations with respect to *debt* are much more homogenous and do not unfairly punish younger individuals. These counterfactuals are also plausible with respect to the mutable feature. Thus, we consider the model in panel (d) as the most explainable according to Def. \ref{def-explainability}.

![Illustration of how CT improves model explainability: (a) conventional training, all mutable; (b) CT, all mutable; (c) conventional, *age* immutable; (d) CT, *age* immutable. The decision boundary is shown in green along with training data colored according to their ground-truth label: $y^+=2$ (orange) and $y^-=1$ (blue). Stars indicate CEs for the target class.](/paper/figures/poc.svg){#fig-poc}

### 2025-03-15 paper/sections/other/main_results.qmd
vamooooos

tbl_mtbl = final_mutability(mtbl_dir; var=["distance"])
df = final_table(res_dir; tbl_mtbl)
inputs = get_table_inputs(df, nothing; backend=Val(:latex))
tabulate_results(inputs; table_type=:tabular, save_name="paper/quarto_ecml/tables/main.tex", wrap_table=false)
tabulate_results(inputs; table_type=:tabular, save_name="paper/preprint/tables/main.tex", wrap_table=false)
```

>>>>

tbl_mtbl = final_mutability(mtbl_dir; var=["distance"])
df = final_table(res_dir; tbl_mtbl)
inputs = get_table_inputs(df, nothing; backend=Val(:latex))
formatters = (
    PrettyTables.ft_round(1,findall(.!contains.(names(df),"Acc"))),
    PrettyTables.ft_round(2,findall(contains.(names(df),"Acc")))
)
for fname in ["paper/quarto_ecml/tables/main.tex","paper/preprint/tables/main.tex"]
    tabulate_results(
        inputs; 
        table_type=:tabular, 
        save_name=fname, 
        wrap_table=false, 
        alignment=[:l,fill(:r,size(df,2)-1)...],
        formatters=formatters
    )
end
```
